Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-08-03 paint-contour-tree.tex
small changes
fixed issue with Claim 8.15
claim 8.15 modified

\begin{claim} \label{clm:weights} There exists a function $\alpha:N(\pruned) \mapsto \RR^+$
with the following properties. (We use $\alpha_p$ to denote the function value.)
Let $A_p = \sum_{q \in \pruned_p} \alpha_q$.
Then, $A_p - 3\sum_{q \in \pruned_p, q \neq p} |q| \leq |H_p| \leq A_p$.
\end{claim}

\begin{proof} We define $\alpha$ inductively, from leaves upwards.
For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider an internal node $p$
with children $q_1, q_2, \ldots, q_k$. The heap $H_p$ is formed by the union
of the $H_{q_i}$ heaps, after deleting certain vertices from them, and potentially
adding some new vertices. The new vertices are potentially those that come
from other nodes that have been pruned off.
The vertices in each $q_i$ as well as $\lambda(q_i)$ are deleted.
Define $\alpha_p$ to be the number of new vertices added to $H_p$.

Thus, we have $|H_p| = \sum_{i \leq k} |H_q| + \alpha_p$
$- \textrm{\# vertices deleted from $H_q$'s}$.
The number of vertices deleted is at most $2\sum_{i \leq k} (|q_i| + |\lambda(q_i)|)$.
There is a factor $2$ because each vertex appears in at most $2$ heaps.

It is evident that $|H_p| \leq \sum_{q \in \pruned_p, q \neq p} \alpha_q = A_p$.
Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} (|q| + |\lambda(q)|)$.
Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(p)|
\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
\end{proof}

@@

\begin{claim} \label{clm:exp-A} For any $q, q' \in \pruned_p$ where $q$
is a parent of $q'$, $A_q \geq (4/3)A_{q'}$.
\end{claim}

\begin{proof} We prove this by induction over the depth of $q$.
When $q$ is a leaf, this is vacuously true. Consider some internal node $q$,
with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
hypothesis to the subshrubs rooted at $q'_i$, we can invoke \Clm{charge}
on each subshrub $\pruned_{q'_i}$. Thus, for all $i$,
$\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
By \Clm{weights}, $|H_{q'_i}| \leq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
to complete the proof.
\end{proof}

@@

we get that $\sum_{q' \in \pruned_{q_i}} |q| \leq A_{q_i}/20$.
Summing this over the children, $\sum_{p \in \pruned, p \neq r} \leq A_r/20$.
By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r|$.

>>>>

\begin{claim} \label{clm:weights} There exists a function $\alpha:N(\pruned) \mapsto \RR^+$
with the following properties. (We use $\alpha_p$ to denote the function value.)
Let $A_p = \sum_{q \in \pruned_p} \alpha_q$.
Then, $A_p - 3\sum_{q \in \pruned_p} |q| \leq |H_p| \leq A_p$.
\end{claim}

\begin{proof} We define $\alpha$ inductively, from leaves upwards.
For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider an internal node $p$
with children $q_1, q_2, \ldots, q_k$. The heap $H_p$ is formed by the union
of the $H_{q_i}$ heaps, after deleting certain vertices from them, and potentially
adding some new vertices. 
The vertices in $p$ and those in the $\lambda(q_i)$ are deleted.
Define $\alpha_p$ to be the number of new vertices added to $H_p$.

Thus, we have $|H_p| = \sum_{i \leq k} |H_{q_i}| + \alpha_p$
$- \textrm{\# vertices deleted from $H_{q_i}$'s}$.
The number of vertices deleted is at most $2(|p|+\sum_{i \leq k} |\lambda(q_i)|)$.
There is a factor $2$ because each vertex appears in at most $2$ heaps.

It is evident that $|H_p| \leq \sum_{q \in \pruned_p} \alpha_q = A_p$.
Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} |\lambda(q)|$ $-2\sum_{q \in \pruned_p} |q|$.
Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(p)|
\leq \sum_{q \in \pruned_p, q \neq p} |q|/100$ $\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
\end{proof}

@@

\begin{claim} \label{clm:exp-A} Fix any path $p \in \pruned \setminus r$.
For any $q, q' \in \pruned_p$ where $q$
is a parent of $q'$, $A_q \geq (5/4)A_{q'}$.
\end{claim}

\begin{proof} We prove this by induction over the depth of $q$.
When $q$ is a leaf, this is vacuously true. Consider some internal node $q$,
with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
hypothesis to the subshrubs rooted at $q'_i$, we can invoke \Clm{charge}
on each subshrub $\pruned_{q'_i}$. Thus, for all $i$,
$\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
By \Clm{weights}, 
$|H_{q'_i}| \geq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
to complete the proof.
\end{proof}

@@

we get that $\sum_{p \in \pruned_{q_i}} |p| \leq A_{q_i}/20$.
Summing this over the children, $\sum_{p \in \pruned, p \neq r} |p| \leq A_r/20$.
By \Clm{weights}, $|H_r| \geq 17A_r/20 - 3|r|$. Since $r$ represented a short path, 
we have the bound $|r| \leq \sqrt{|H_r|}/100$. Thus, $|H_r| = \Omega(A_r)$.
